Concurrent Programming in Harmony

### 2022-01-11 book/paper.tex
Latest

\begin{figure}
\begin{code}
<{:input:../code/linkedlist.hny}>
\end{code}
\caption{\harmonylink{code/linkedlist.hny} Implementation of a set of integers using a linked list with fine-grained locking}
\label{fig:linkedlist}
\end{figure}

@@

\autoref{fig:linkedlist} implements a concurrent set object using
an ordered linked list of integers without duplicates.
The list has two dummy ``book-end'' nodes with values <{-inf}> and
<{inf}> (similar to the Python \texttt{math.inf} constant).
An invariant of the algorithm is that at any point in time the
list is ``valid,'' starting with a <{-inf}> node and ending
with an <{inf}> node.

\index{hand-over-hand locking}%
Each node has a lock, a value, and <{next}>, a pointer to the next node
(which is <{None}> for the <{inf}> node).
The <{_find(lst, v)}> helper method first
finds and locks two consecutive nodes <{before}> and <{after}>
such that <{before->data.value < v <= after->data.value}>.
It does so by performing something called \emph{hand-over-hand locking}.
It first locks the first node, which is the <{-inf}> node.
Then, iteratively, it obtains a lock on the next node and release the
lock on the last one, and so on, similar to climbing a rope
hand-over-hand.
Using \texttt{\_find}, the <{insert}>, <{remove}>, and \texttt{contains}
methods are fairly straightforward to implement.

\begin{comment}
\glsadd{linearizable}%
Like the queue in \autoref{fig:queues}(b), the implementation of the list is
\emph{linearizable}~\cite{HW90},
\index{linearizable}%
a strong notion of consistency that
makes it appear as if each of the operations executes atomically at
some point between their invocation and return.
Determining if an implementation of a concurrent data structure
is linearizable involves finding what are known as the
\emph{linearization points}
\index{linearization point}%
of the operations in an execution.  These are the unique
points in time at which an operation appears to execute atomically.
The linearization point for the <{insert}> operation coincides exactly with the update
of the <{before.next}> pointer.
The linearization point of a \texttt{contains} method
execution depends on whether the value is found or not.
If found, it coincides with retrieving the pointer to the node
that has the value.
If not found, it coincides with retrieving the pointer to the
<{inf}> node.
\end{comment}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
Boolean & <{False}>, <{True}> \\
\hline
Integer & <{-inf, ..., -2, -1, 0, 1, 2, ..., inf}> \\
\hline
String & <{"example"}>, <{.example}> \\
\hline
Program Counter & (method names are program counter constants) \\
\hline
Dictionary & <{{ .account: 12345, .valid: False }}> \\
\hline
Set & <{{}, { 1, 2, 3 }, { False, .id, 3 }}> \\
\hline
Address & <{?lock, ?flags[2], None}> \\
\hline
Context & (generated by <{stop}> expression) \\
\hline
\end{tabular}
\vspace{1em}

@@

The integer type supports any whole number, as well as $-\mathtt{inf}$
and $\mathtt{inf}$.  In the Python-based model checker, integers are
infinite precision.  In the C-based model checker, integers are implemented
by two's complement 61-bit words, and $-\mathtt{inf}$ is represented by
the minimal integer and $\mathtt{inf}$ is represented by the maximal integer.

>>>>

\begin{figure}
\begin{code}
<{:input:../code/linkedlist.hny}>
\end{code}
\caption{\harmonylink{code/linkedlist.hny} Implementation of a set of values using a linked list with fine-grained locking}
\label{fig:linkedlist}
\end{figure}

@@

\autoref{fig:linkedlist} implements a concurrent set object using
an ordered linked list without duplicates.
The list has two dummy ``book-end'' nodes with values <{(-1, None)}> and
<{(1, None)}>.  A value <{v}> is stored as <{(0, v)}>---note that for
any value <{v}>, <{(-1, None) < (0, v) < (1, None)}>.
An invariant of the algorithm is that at any point in time the
list is ``valid,'' starting with a <{(-1, None)}> node and ending
with an <{(1, None)}> node.

\index{hand-over-hand locking}%
Each node has a lock, a value, and <{next}>, a pointer to the next node
(which is <{None}> for the <{(1, None)}> node to mark the end of the list).
The <{_find(lst, v)}> helper method first
finds and locks two consecutive nodes <{before}> and <{after}>
such that <{before->data.value < (0, v) <= after->data.value}>.
It does so by performing something called \emph{hand-over-hand locking}.
It first locks the first node, which is the <{(-1, None)}> node.
Then, iteratively, it obtains a lock on the next node and release the
lock on the last one, and so on, similar to climbing a rope
hand-over-hand.
Using \texttt{\_find}, the <{insert}>, <{remove}>, and \texttt{contains}
methods are fairly straightforward to implement.

\begin{comment}
\glsadd{linearizable}%
Like the queue in \autoref{fig:queues}(b), the implementation of the list is
\emph{linearizable}~\cite{HW90},
\index{linearizable}%
a strong notion of consistency that
makes it appear as if each of the operations executes atomically at
some point between their invocation and return.
Determining if an implementation of a concurrent data structure
is linearizable involves finding what are known as the
\emph{linearization points}
\index{linearization point}%
of the operations in an execution.  These are the unique
points in time at which an operation appears to execute atomically.
The linearization point for the <{insert}> operation coincides exactly with the update
of the <{before.next}> pointer.
The linearization point of a \texttt{contains} method
execution depends on whether the value is found or not.
If found, it coincides with retrieving the pointer to the node
that has the value.
If not found, it coincides with retrieving the pointer to the
<{(1, None)}> node.
\end{comment}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
Boolean & <{False}>, <{True}> \\
\hline
Integer & <{..., -2, -1, 0, 1, 2, ...}> \\
\hline
String & <{"example"}>, <{.example}> \\
\hline
Program Counter & (method names are program counter constants) \\
\hline
Dictionary & <{{ .account: 12345, .valid: False }}> \\
\hline
Set & <{{}, { 1, 2, 3 }, { False, .id, 3 }}> \\
\hline
Address & <{?lock, ?flags[2], None}> \\
\hline
Context & (generated by <{stop}> expression) \\
\hline
\end{tabular}
\vspace{1em}

@@

The integer type supports any whole number.
In the C-based model checker, integers are currently implemented
by two's complement 61-bit words.  The model checker checks for
overflow on various operations.
